Rectangulotopes

### 2024-03-25 draft/rectangulotopes.tex
figures rectangulation lattices
updates

% polytopes
\newcommand{\Asso}{\polytope{A}\mathsf{sso}} % associahedron
% rectangulotopes
\newcommand{\WRP}{\polytope{WR}} % weak rectangulation polytope
\newcommand{\SRP}{\polytope{SR}} % strong rectangulation polytope
% shard polytopes
\newcommand{\SP}{\polytope{SP}}

@@

\title{Rectangulation polytopes}

@@

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We describe realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles. They are defined as quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations. We give  elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedra.
\end{abstract}

@@

A striking yet elementary example of this idea is the \defn{permutahedron} $\polytope{P} (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \pi^{-1}(i)\cdot \mathbf{e}_i$, for all permutations $\pi\in\f{S}_n$. The facets of $\polytope{P} (n)$ are indexed by the nonempty and nontrivial subsets of $[n]$, while its skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.

The \defn{associahedron} is an ubiquitous polytope, first defined by Tamari~\cite{}, then studied by Stasheff in a topological context~\cite{}, that has since then found applications in various areas of mathematics~\cite{}. It was first defined as a combinatorial object, before geometric realizations were shown to exist~\cite{}.

@@

\begin{theorem}
  Let $\Asso (n)$ be the convex hull of the points
  $\sum_{i\in [n]} w_i\cdot \mathbf{e}_i$, where
  \[
  w_i := \ell(T, i)\cdot r(T,i),
  \]
  for all binary trees $T$ on $n$ internal nodes, where $\ell(T,i)$ and $r(T,i)$ denote respectively the number of leaves in the left and right subtrees of $i$ in $T$.
  Then $\Asso (n)$ is a realization of the $(n-1)$-dimensional associahedron.
\end{theorem}

\begin{itemize}
\item \defn{Generalized (deformed) permutahedra}
\item \defn{Quotientopes}
\item Quotientopes of rectangulations, natural successors to associahedra; Definitions
\item We give a Loday-style description of two classes of rectangulation polytopes
\item This intro states the main results; proofs are in the following sections.
\end{itemize}

@@

@@

\subsection{Weak rectangulation polytopes}

We are now ready to give our first main result on the realizations of weak rectangulation polytopes.

\begin{theorem}
  Let $\WRP (n)$ be the convex hull of the points
  $\sum_{i\in [n]} (\loday{w}_i - \antiloday{w}_i)\cdot \mathbf{e}_i$,
  where
  \[
  \begin{split}
    \loday{w}_i := & h(T(R), i)\cdot v(T(R),i) \\
    \antiloday{w}_i := & h(S(R), i)\cdot v(S(R),i),
  \end{split}
  \]
  for all weak rectangulations $R$ on $n$ rectangles, where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
  Then $\WRP (n)$ is a realization of the $(n-1)$-dimensional weak rectangulation polytope.
\end{theorem}

\subsection{Strong rectangulation polytopes}

Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$. We denote by $i|j$ and $\frac{\ j\ }i$ the situations where the rectangle labeled $i$ touches the rectangle labeled $j$ from the left and from below in $R$, respectively.

\begin{theorem}
  Let $\SRP (n)$ be the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}_{i,j} - \yang{w}_{i,j})\cdot (\mathbf{e}_i - \mathbf{e}_j),
  \]
  with
  \[
  \yin{w}_{i,j} := 
  \begin{cases}
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) & \text{ if neither } i|j \text{ nor } \frac{\ j\ }i, \\
    0 & \text{ otherwise.}
  \end{cases}
  \]
  and
    \[
  \yang{w}_{i,j} := 
  \begin{cases}
    v(S(R), j) \cdot ch (T(R), S(R), i, j)\cdot v(T(R), i) & \text{ if neither } j|i \text{ nor } \frac{\ i\ }j, \\
    0 & \text{ otherwise.}
  \end{cases}
  \]

  \jean{Double check the formula.}
  for all strong rectangulations $R$ on $n$ rectangles, where:
  \begin{itemize}
  \item $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$,
  \item $ch (T,S,i,j)$ and $cv (T,S,i,j)$ denote respectively the number of common leaves of the horizontal (vertical) subtree of $i$ in $T$ and the horizontal (vertical) subtree of $j$ in $S$
  \end{itemize}
Then $\SRP (n)$ is a realization of the $(n-1)$-dimensional strong rectangulation polytope.  
\end{theorem}

@@

\section{Weak rectangulation polytopes}
\label{sec:weakRectangulationPolytopes}

@@

We call this congruence the \defn{Baxter congruence}.
The \defn{weak rectangulation polytope} $\WRP(n)$ is the quotientope of the Baxter congruence $\baxtereq$ on the weak Bruhat order on $\f{S}_n$.

@@

\begin{lemma}
  The weak rectangulation polytope $\WRP(n)$ is realized by the Minkowski sum of all Loday and anti-Loday shard polytopes. In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

\subsection{Loday-type coordinates for weak rectangulation polytopes}

@@

\section{Strong rectangulation polytopes}
\label{sec:strongRectangulationPolytopes}

@@

We call the congruence $\strongeq$ the \defn{strong rectangulation congruence}.
The \defn{strong rectangulation polytope} $\SRP(n)$ is the quotientope of the strong rectangulation congruence $\strongeq$.

@@

\begin{lemma}
  \label{lem:strongsum}
  The strong rectangulation polytope $\SRP(n)$ is realized by the Minkowski sum of all shard polytopes of yin and yang arcs.
\end{lemma}

@@

\subsection{Loday-type coordinates for strong rectangulation polytopes}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\SRP(n)$ associated with $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $A$ such that $\mathbf{e}_i-\mathbf{e}_j$
  is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
  \yin{w}_{i,j} = 
  \begin{cases}
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) & \text{ if neither } i|j \text{ nor } \frac{\ j\ }i \\
    0 & \text{ otherwise }
  \end{cases}
  \]
\end{lemma}
\begin{proof}
  Let $\pi$ be the 2-clumped permutation corresponding to $R$, hence the minimal element of $\mathcal{L}(P_s(R))$ in the weak Bruhat order.
  From Lemma~\ref{lem:yinminmax}, the set of yin arcs $A$ such that $\SP(A)$ contributes to $\mathbf{e}_i-\mathbf{e}_j$
  are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that:
  \begin{itemize}
  \item $i = \arg\max_{k\in I} \pi^{-1}(k)$,
  \item $j = \arg\min_{k\in J} \pi^{-1}(k)$, and
  \item $i,j$ is an inversion in $\pi$, hence $\pi^{-1}(i)>\pi^{-1}(j)$.
  \end{itemize}
  We observe that provided the third condition holds, the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the product of the
  number of leaves of the subtree of $T(R)$ rooted at $i$, the number of common leaves to the vertical subtrees of $T(R)$ and $S(R)$ rooted
  at $i$ and $j$, respectively, and the number of leaves of the subtree of $S(R)$ rooted at $j$.

>>>>

% polytopes
\newcommand{\Asso}{\polytope{A}\mathsf{sso}} % associahedron
% rectangulotopes
\newcommand{\WRP}{\polytope{WR}} % weak rectangulotope
\newcommand{\SRP}{\polytope{SR}} % strong rectangulotope
% shard polytopes
\newcommand{\SP}{\polytope{SP}}

@@

%\title{Rectangulation polytopes}
\title{Rectangulotopes}

@@

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We describe realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles. They are defined as quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations. We give elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedron.
\end{abstract}

@@

An elementary yet striking example of this idea is the \defn{permutahedron} $\polytope{P} (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \pi^{-1}(i)\cdot \mathbf{e}_i$, for all permutations $\pi\in\f{S}_n$. The facets of $\polytope{P} (n)$ are indexed by the nonempty and nontrivial subsets of $[n]$, while its skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.

The \defn{associahedron} is an ubiquitous polytope, first defined by Tamari~\cite{T51}, then studied by Stasheff in a topological context~\cite{S63}, that has since then found applications in various areas of mathematics, including topology~\cite{}, operads~\cite{}, cluster algebras~\cite{}, combinatorial Hopf algebras~\cite{}, physics~\cite{}, etc. It was first defined as a combinatorial object, before geometric realizations were shown to exist~\cite{}.

@@

\begin{theorem}
  \label{thm:loday}
  Let $\Asso (n)$ be the convex hull of the points
  $\sum_{i\in [n]} w_i\cdot \mathbf{e}_i$, where
  \[
  w_i := \ell(T, i)\cdot r(T,i),
  \]
  for all binary trees $T$ on $n$ internal nodes, where $\ell(T,i)$ and $r(T,i)$ denote respectively the number of leaves in the left and right subtrees of $i$ in $T$.
  Then $\Asso (n)$ is a realization of the $(n-1)$-dimensional associahedron.
\end{theorem}

A natural object associated with the associahedron is the \defn{Tamari lattice}, a lattice whose cover graph is isomorphic to the skeleton of the associahedron. The Tamari lattice is known to be the quotient of the weak Bruhat order by a \defn{lattice congruence} relation called the \defn{sylvester congruence}. Pilaud and Santos~\cite{}, answering a question of Reading~\cite{}, proved that with \emph{every} lattice congruence of the weak order, one can associate a polytope whose skeleton is the cover graph of the lattice quotient, and more generally whose normal fan is the fan defined by gluing the cones of the \defn{braid fan} (the type $A$ Coxeter arrangement) that belong to the same congruence class. They called such polytopes \defn{quotientopes}.

The goal of this paper is to initiate the study of quotientopes of lattice congruences whose classes are in bijections with \defn{rectangulations}, defined as decompositions of a square into finitely many axis-aligned rectangles.

Those lattice congruences are the \defn{Baxter congruence} on one hand, studied by Law and Reading~\cite{}, and whose classes are one-to-one with \defn{weak rectangulations} (or \defn{diagonal rectangulations}), and the \defn{strong rectangulation congruence}, whose classes are in bijection with so-called \defn{strong rectangulations}. The latter is mentioned by Reading~\cite{} (seee for instance in Theorem 1.2, and Example 4.11), and studied more recently by Asinowski et al.~\cite{}.

One of our main results is an elementary description of the vertex coordinates of a realization of both families of quotientopes, in the spirit of Theorem~\ref{thm:loday}.
The quotientopes corresponding to the Baxter congruence will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$. similarly, the polytopes corresponding to the strong rectangulation congruence will be called the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.

In this introduction, we give a minimal set of definition to be able to completely state our results on the vertex description of the weak and strong rectangulotopes. Section~\ref{sec:quotientopes} is dedicated to the background on quotientopes and their realizations as Minkowski sums of \defn{shard polytopes}, due to Padrol, Pilaud, and Ritter~\cite{}. Section~\ref{sec:wr} details the case of weak rectangulations, while Section~\ref{sec:sr} deals with strong rectangulations.

@@

Law and Reading~\cite{} defined a flip graph on weak rectangulations.

\begin{lemma}
  The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$.
\end{lemma}

Meehan~\cite{} (See also Asinowski et al.~\cite{}) describe the flip graph on strong rectangulations.

\begin{lemma}
  The skeleton of $\SRP(n)$ is isomorphic to the flip graph of strong rectangulations of size $n$.
\end{lemma}

@@

\subsection{Weak rectangulotopes}

We are now ready to give our first main result on the realizations of weak rectangulotopes.

\begin{theorem}
  Let $\WRP (n)$ be the convex hull of the points
  $\sum_{i\in [n]} (\loday{w}_i - \antiloday{w}_i)\cdot \mathbf{e}_i$,
  where
  \[
  \begin{split}
    \loday{w}_i := & h(T, i)\cdot v(T,i) \\
    \antiloday{w}_i := & h(S, i)\cdot v(S,i),
  \end{split}
  \]
  for all weak rectangulations $R$ on $n$ rectangles, where $T:=T(R)$, $S:=S(R)$,
  and $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
  Then $\WRP (n)$ is a realization of the $(n-1)$-dimensional weak rectangulotope.
\end{theorem}

\subsection{Strong rectangulotopes}

Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$. We denote by %$i|j$ and $\frac{\ j\ }i$
$i\backslash j$ the situation where the rectangle labeled $i$ touches the rectangle labeled $j$ either from the left or from below in $R$. %, respectively.

\begin{theorem}
  Let $\SRP (n)$ be the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}_{i,j} - \yang{w}_{i,j})\cdot (\mathbf{e}_i - \mathbf{e}_j),
  \]
  with
  \[
  \begin{split}
    \yin{w}_{i,j} := & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot [ \neg i\backslash j ] \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
    \yang{w}_{i,j} := & v(S, j) \cdot ch (T, S, i, j)\cdot v(T, i) \cdot [ \neg j\backslash i ]
  \end{split}
  \]

%  \begin{cases}
%    h(T(R), i) \cdot cv (T, S, i, j)\cdot h(S, j) & \text{ if neither } i|j \text{ nor } \frac{\ j\ }i, \\
%    0 & \text{ otherwise.}
%  \end{cases}

  %\begin{cases}
  %  v(S, j) \cdot ch (T, S, i, j)\cdot v(T, i) & \text{ if neither } j|i \text{ nor } \frac{\ i\ }j, \\
  %  0 & \text{ otherwise.}
  %\end{cases}

  \jean{Double check the formula.}
  for all strong rectangulations $R$ on $n$ rectangles, where:
  \begin{itemize}
  \item $T := T(R)$ and $S:=S(R)$,
  \item $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$,
  \item $ch (T,S,i,j)$ and $cv (T,S,i,j)$ denote respectively the number of common leaves of the horizontal (vertical) subtree of $i$ in $T$ and the horizontal (vertical) subtree of $j$ in $S$
  \end{itemize}
Then $\SRP (n)$ is a realization of the $(n-1)$-dimensional strong rectangulotope.  
\end{theorem}

@@

\section{Weak rectangulotopes}
\label{sec:wr}

@@

We call this congruence the \defn{Baxter congruence}.
The \defn{weak rectangulotope} $\WRP(n)$ is the quotientope of the Baxter congruence $\baxtereq$ on the weak Bruhat order on $\f{S}_n$.

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all Loday and anti-Loday shard polytopes. In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

\subsection{Loday-type coordinates for weak rectangulotopes}

@@

\section{Strong rectangulotopes}
\label{sec:sr}

@@

We call the congruence $\strongeq$ the \defn{strong rectangulation congruence}.
The \defn{strong rectangulotope} $\SRP(n)$ is the quotientope of the strong rectangulation congruence $\strongeq$.

@@

\begin{lemma}
  \label{lem:strongsum}
  The strong rectangulotope $\SRP(n)$ is realized by the Minkowski sum of all shard polytopes of yin and yang arcs.
\end{lemma}

@@

\subsection{Loday-type coordinates for strong rectangulotopes}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\SRP(n)$ associated with $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $A$ such that $\mathbf{e}_i-\mathbf{e}_j$
  is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) 
  \]
  if neither $i|j$ nor $\frac{\ j\ }i$, and $0$ otherwise.
\end{lemma}
\begin{proof}
  Let $\pi$ be the 2-clumped permutation corresponding to $R$, hence the minimal element of $\mathcal{L}(P_s(R))$ in the weak Bruhat order.
  From Lemma~\ref{lem:yinminmax}, the set of yin arcs $A$ such that $\SP(A)$ contributes to $\mathbf{e}_i-\mathbf{e}_j$
  are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that:
  \begin{itemize}
  \item $i = \arg\max_{k\in I} \pi^{-1}(k)$,
  \item $j = \arg\min_{k\in J} \pi^{-1}(k)$, and
  \item $i,j$ is an inversion in $\pi$, hence $\pi^{-1}(i)>\pi^{-1}(j)$.
  \end{itemize}
  We observe that provided the third condition holds, the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the product of the
  number of leaves of the subtree of $T(R)$ rooted at $i$, the number of common leaves to the vertical subtrees of $T(R)$ and $S(R)$ rooted
  at $i$ and $j$, respectively, and the number of leaves of the subtree of $S(R)$ rooted at $j$.
